---
title: "TousAntiCovid"
author: "FD"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    self_contained: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(global.par = TRUE)
```

# Data

-  TousAntiCovid: [open data, official source](https://www.data.gouv.fr/fr/datasets/metriques-dutilisation-de-lapplication-tousanticovid/)
-  Epidemic data: [JHU data](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series), because recorded as function of date of the result's announcement. 


```{r, results="hide"}
# Load and format data

# 1) TAC
# Data Source : https://www.data.gouv.fr/fr/datasets/metriques-dutilisation-de-lapplication-tousanticovid/
# URL of the data on data.gouv.fr
URL <- "https://www.data.gouv.fr/fr/datasets/r/0e2168ec-24c7-4b49-a900-c7dd12f8e88c"

# Download and read data
today <- Sys.Date() # today's date
dataFile <- paste0("data/TAC_", today, ".csv") # name file with today's date
download.file(URL, dataFile) # download file from repo
dat <- read.csv(dataFile)

## Description from the website (2020-12-23)
## Le présent jeu de données informe pour chaque jour depuis le lancement de l'application le 2 juin 2020 :
##
## Total cumulé du nombre d'applications enregistrées moins le nombre de désenregistrements.
## Total cumulé d’utilisateurs notifiés par l’application : le nombre d’utilisateurs notifiés par l’application comme contacts à risque suite à une exposition à la COVID-19, depuis le 2 juin 2020.
## Total cumulé d’utilisateurs se déclarant comme des cas de COVID-19 par jour : le nombre d’utilisateurs qui se sont déclarés comme des cas de COVID-19 dans l’application, depuis le 2 juin 2020.

# Check data
# Time differences between dates
unique(diff(as.Date(dat$date))) # -> there are some gaps in the data

# New columns with delta between dates
dat$diffDate <- c(NA, diff(as.Date(dat$date))) # Difference in days between data points
# for the other columns, divide by the time difference to have differences per day
# (otherwise data not comparable when different time differences)
dat$diffRegisterPerDay <- c(NA, diff(dat$register) / dat$diffDate[-1]) 
dat$diffQRPerDay <- c(NA, diff(dat$qrCode) / dat$diffDate[-1])
dat$diffNotifPerDay <- c(NA, diff(dat$notified) / dat$diffDate[-1])

# Create new dataset without gaps in the dates
# New dates, without gaps
rangeDates <- range(as.Date(dat$date))
allDates <- seq(rangeDates[1], rangeDates[2], by="days")
dat.temp <- as.data.frame(as.character(allDates))
names(dat.temp) <- "date"

# Merge the datasets by date 
tmp <- merge(dat, dat.temp, by = "date", all = TRUE)
# note: this adds NA at the dates where there were no data; 
# these dates are put at the end of the table

# Sort the new dataset by date
sortedDateIndex <- sort(as.character(tmp$date), index.return = TRUE)$ix
dat.all <- tmp[sortedDateIndex, ]

# Add day of the week
dat.all$weekDay <- weekdays(as.Date(dat.all$date))

# Load function to compute sliding average
source("commonFuncs.R")

# Remove negative values
dat.all$diffNotifPerDay[dat.all$diffNotifPerDay < 0 ] <- NA

# Compute the average values, 1-week window, focal point in the middle
dat.all$diffRegisterPerDay.WeekAve <- sliding.ave(dat.all$diffRegisterPerDay)
dat.all$diffQRPerDay.WeekAve <- sliding.ave(dat.all$diffQRPerDay)
dat.all$diffNotifPerDay.WeekAve <- sliding.ave(dat.all$diffNotifPerDay)
dat.all$register.WeekAve <- sliding.ave(dat.all$register)
dat.all$date <- as.Date(dat.all$date)

```

```{r}
# 2) Data from JHU
# The date in these data is the date of the result's annoucement
URL <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
# Download and read data
dataFile <- paste0("data/JHU_cases_all_", today, ".csv") # name file with today's date
download.file(URL, dataFile) # download file from repo
datJHU <- read.csv(dataFile, header = TRUE, check.names = FALSE)
# Extract data about France
datJHU.France <- datJHU[datJHU$`Country/Region` == "France", ]

# Collapse data for all France (including DOM TOM)
datJHU.France.agg <- apply(datJHU.France[, -(1:4)], 2, sum)
# Need to reformat the data: have one column for dates and one column for values
datJHUFr <- data.frame(matrix(0, ncol = 2, nrow = length(datJHU.France.agg)))
datJHUFr$date <- as.Date(names(datJHU.France.agg), format = "%m/%d/%y")
datJHUFr$cumP <- unname(datJHU.France.agg)
datJHUFr <- datJHUFr[, -(1:2)]
# Get week day
datJHUFr$weekDay <- weekdays(as.Date(datJHUFr$date))
# Extract data per day (JHU provides cumulative values)
datJHUFr$P <- c(datJHUFr[1, "cumP"], diff(datJHUFr$cumP))
# Add the sliding window value to the JHU data
datJHUFr$P.WeekAve <- sliding.ave(datJHUFr$P, winwdt = 7, pos = 4, na.rm = TRUE)
```

```{r}
# 3) French data from data.gouv.fr, which JHU uses as source
URL <- "https://www.data.gouv.fr/en/datasets/r/d3a98a30-893f-47f7-96c5-2f4bcaaa0d71"
# Download and read data
dataFile <- paste0("data/FR_cases_all_", today, ".csv") # name file with today's date
download.file(URL, dataFile) # download file from repo
datFR <- read.csv(dataFile, header = TRUE, check.names = FALSE)
datFR$date <- as.Date(datFR$date)

# Get week day
datFR$weekDay <- weekdays(as.Date(datFR$date))
# Extract data per day (JHU provides cumulative values)
datFR$P <- c(datFR[1, "total_cas_confirmes"], diff(datFR$total_cas_confirmes))
# Add the sliding window value to the JHU data
datFR$P.WeekAve <- sliding.ave(datFR$P, winwdt = 7, pos = 4, na.rm = TRUE)

```


```{r}
# Merge the datasets
dat.withEpi <- merge(dat.all, datFR, by = "date")
names(dat.withEpi)[9] <- "weekDay"
```


Données TAC jusqu'au `r format(max(dat.all$date), "%a %d %b %Y")` (inclus)   
Données épidémio (JHU) jusqu'au `r format(max(datJHUFr$date), "%a %d %b %Y")` (inclus).
Données épidémio (data.gouv.fr) jusqu'au `r format(max(as.Date(datFR$date)), "%a %d %b %Y")` (inclus).


## Quelques dates clés

- StopCovid est devenu TousAntiCovid vers le 22 octobre 2020 https://www.gouvernement.fr/partage/11818-conference-de-presse-sur-l-application-des-mesures-contre-la-covid-19
```{r  include = FALSE}
date_TAC <- "2020-10-22"
```

- Une campagne de publicité a été lancée à la fin novembre / début décembre 2020 (avec envoi de SMS à la population)  

- Les critères de détection de contact ont changé le 27-28 novembre 2020 (changement distance et durée). [Source](https://www.service-public.fr/particuliers/actualites/A14486)  
```{r  include = FALSE}
date_criteres <- "2020-11-28"
```

- A partir du 16 décembre 2020, le QR code est plus facile à obtenir (envoi SMS). [Source](https://twitter.com/TousAntiCovid/status/1349013051387871235?s=20).
```{r  include = FALSE}
date_sms <- "2020-12-16"
```

- Nouveau critère encore de détection de contacts vers le 9 mars 2021.  
[Source](https://twitter.com/TousAntiCovid/status/1369668376709193730?s=20).  
```{r include = FALSE}
date_criteres2 <- "2021-03-09"
```


# Plots

```{r, results="hide", fig.show="hide"}
# Load plot settings
source("commonPlot.R")
```

## TAC, registrations

### Number of registered users, over time

```{r}
plot(as.Date(dat.all$date), dat.all$register, 
     xlab = "date", 
     ylab = "Registrations", 
     pch = 16, col = cols[dat.all$weekDay])
```

### Changes in number of registered users, per day  
(NB: when there is a gap in the data, the difference in number of users is divided by the length of the time gap)  
```{r}
plot(as.Date(dat.all$date), dat.all$diffRegisterPerDay, log = "",      
     xlab = "date", 
     ylab = "delta registrations per day", 
     pch = 16, col = cols[dat.all$weekDay], 
     ylim = c(1, 7*10^5), 
     yaxs = "i")
legend(as.Date("2020-07-01"), y = 500000, legend = names(cols), col = cols, pch = 16, box.lwd = 0)
abline(v = as.Date("2020-10-22"))

```

The vertical line corresponds to the switch from StopCovid to TousAntiCovid. 
There was an advertisement campaign by the end of November/beginning of December, with SMS sent to the population. 

## TAC, QRcodes

### QR codes scanned every day

```{r}
plot(as.Date(dat.all$date), dat.all$diffQRPerDay, log = "",      
     xlab = "date", 
     ylab = "delta QR per day", 
     pch = 16, col = cols[dat.all$weekDay])
lines(dat.all$date, dat.all$diffQRPerDay.WeekAve, col = colAve)
legend(as.Date("2020-07-01"), y = 2000, legend = c(names(cols),  "7-day average"), col = c(cols, colAve), pch = c(rep(16, 7), -1), box.lwd = 0, lwd = c(rep(0, 7), 1))
```

The minima are on Sundays

## TAC, notifications

### Notifications sent every day  
```{r}
plot(as.Date(dat.all$date), dat.all$diffNotifPerDay, log = "",      
     xlab = "date", 
     ylab = "delta notifications per day", 
     pch = 16, col = cols[dat.all$weekDay])

lines(dat.all$date, dat.all$diffNotifPerDay.WeekAve, col = colAve)

legend(as.Date("2020-07-01"), y = 1000, legend = c(names(cols),  "7-day average"), col = c(cols, colAve), pch = c(rep(16, 7), -1), box.lwd = 0, lwd = c(rep(0, 7), 1))
```

The minima are on Mondays  
-> There seems to be a one-day difference between scanned QR codes and notifications sent. 
This was already noticed by F. Lesueur [here](https://flesueur.medium.com/réalignement-temporel-des-chiffres-tousanticovid-covid-7bda2f0bf1c1). 

### Notifications per QR code

1) same day   
```{r}
plot(as.Date(dat.all$date), dat.all$diffNotifPerDay / dat.all$diffQRPerDay, log = "",      
     xlab = "date", 
     ylab = "notifs per QRcode", 
     pch = 16, col = cols[dat.all$weekDay], main = "raw data, same day")
```

2) one-day difference
```{r}
plot(as.Date(dat.all$date), dat.all$diffNotifPerDay / (c(NA, dat.all$diffQRPerDay[-nrow(dat.all)])), log = "",      
     xlab = "date", 
     ylab = "notifs per QR code", 
     pch = 16, col = cols[dat.all$weekDay], main = "raw data, one-day delay in notifications")
```


3) same day, 7-day average   
```{r}
plot(as.Date(dat.all$date), dat.all$diffNotifPerDay.WeekAve / dat.all$diffQRPerDay.WeekAve, log = "",      
     xlab = "date", 
     ylab = "notifs per QRcode", 
     pch = 16, col = cols[dat.all$weekDay], main = "7-day ave, same day")
```

4) one-day difference, 7-day average
```{r}
par(mar = rep(5, 4))
colNotifsCas <- "#045a8d"
plot(as.Date(dat.all$date), dat.all$diffNotifPerDay.WeekAve / (c(NA, dat.all$diffQRPerDay.WeekAve[-nrow(dat.all)])), log = "",      
     xlab = "", 
     ylab = "nb de notifs par QR code", 
     pch = 16, col = colNotifsCas, 
     yaxs = "i", frame.plot = FALSE)

title("Nb notifications par QR code, moy. 7j", line = 2.5)

# Annotate with dates
abline(v = as.Date(date_TAC), lty = 2)
cexAnnotate <- 0.8
mtext("StopCovid -> 
TousAntiCovid", side = 3, at = as.Date(date_TAC), adj = 1, cex = cexAnnotate)

abline(v = as.Date(date_sms), lty = 2)
mtext("envoi SMS pour 
QR code", side = 3, at = as.Date(date_sms), adj = 0, cex = cexAnnotate)

abline(v = as.Date(date_criteres), lty = 2)
mtext("nvx criteres
detection", side = 3, at = as.Date(date_criteres), adj = 0.7, cex = cexAnnotate)


abline(v = as.Date(date_criteres2), lty = 2)
mtext("nvx criteres
detection", side = 3, at = as.Date(date_criteres2), adj = 0.5, cex = cexAnnotate)

mtext("Date", side = 1, line = 1.5)

for(i in seq(0.5, 1.5, by = 0.5)){
  abline(h = i, lty = 1, col = gray(0.8), lwd = 0.4)
}
mtext(paste0("Mise à jour : ", Sys.Date(), "
Données : https://www.data.gouv.fr/fr/datasets/metriques-dutilisation-de-lapplication-tousanticovid/ pour TAC"), cex = 0.55, col = gray(0.3), side = 1, line = 3.25, adj = 0, )

axis(4)
```

## Epidemic data

JHU data, date is date at which result was communicated (and not sampling date).

```{r}
plot(dat.withEpi$date, dat.withEpi$P, col = cols[dat.withEpi$weekDay], pch = 16, 
     xlab = "date", ylab = "nb of cases")
lines(dat.withEpi$date, dat.withEpi$P.WeekAve, col = colAve)
legend(as.Date("2020-07-01"), y = 65000, legend = c(names(cols),  "7-day average"), col = c(cols, colAve), pch = c(rep(16, 7), -1), box.lwd = 0, lwd = c(rep(0, 7), 1))
```

## Active users

Not all users are active; some people have downloaded the app but do not turn on bluetooth; some may just have downloaded the app for their "attestations".  
To estimate the number of active users, we will compare the proportion of registered users in the whole population to the proportion of scanned QR codes among positive test results. 

```{r}
par(xpd = TRUE)
# Approximate size of the French population
par(mar = c(5, 5, 5, 5))
par(mgp = c(3, 0.75, 0))
popFR <- 67*10^6

# Registered users, as fraction of the total popuation
propUtil.Raw <- dat.withEpi$register / popFR
propUtil <- dat.withEpi$register.WeekAve / popFR

# Active users, averaged over a week
propUtilActif <- dat.withEpi$diffQRPerDay.WeekAve / dat.withEpi$P.WeekAve

colActif <- rgb(0.4, 0.7, 0)
colRegis <- gray(0.4)
cexPoints <- 0.8 # Size of the dots
ymax <- ceiling(10*max(propUtil.Raw, na.rm = TRUE))/10

plot(dat.withEpi$date, propUtilActif, 
     xlab = "", ylab = "Proportions d'utilisateurs", ylim = c(0, ymax), 
     pch = 16, col = colActif, 
       main = "Fraction de la population utilisant TAC", 
     yaxs = "i", frame.plot = FALSE, cex = cexPoints)
mtext("Date", side = 1, line = 1.5)

# Add horizontal lines to read the values
par(xpd = FALSE)
for(i in seq(0.05, ymax, by = 0.05)){
  abline(h = i, lty = 1, col = gray(0.8), lwd = 0.4)
}

# Add points for raw users
points(dat.withEpi$date, propUtil.Raw, pch = 16, col = colRegis, cex = cexPoints)

legend("topleft", col = c(gray(0.4), rgb(0.4, 0.7, 0)), pch = 16, 
       legend = c("Enregistrés (brut)", "Actifs (moy. 7j)"), 
       bty = "o", pt.cex = cexPoints, box.lwd = 0)

axis(4)
par(xpd = FALSE)

# Annotate with dates
abline(v = as.Date(date_TAC), lty = 2)
abline(v = as.Date(date_sms), lty = 2)
cexAnnotate <- 0.8
dx <- 0.3
mtext("StopCovid -> 
TousAntiCovid", side = 3, at = as.Date(date_TAC), adj = 1-dx, cex = cexAnnotate)

mtext("envoi SMS pour 
QR code", side = 3, at = as.Date(date_sms), adj = dx, cex = cexAnnotate)

mtext(paste0("Mise à jour : ", Sys.Date(), "
Données : https://www.data.gouv.fr/fr/datasets/metriques-dutilisation-de-lapplication-tousanticovid/ pour TAC, 
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series pour nb cas en date d'annonce"), cex = 0.55, col = gray(0.3), side = 1, line = 3.25, adj = 0)
```

## App settings

We want to see if the conditions to detect a contact have changed over time. If the app remains the same, the proportion of notified cases per QRcode should be a constant multiple of the proportion of app users. 
Notifications are sent to contacts of the 14? last days (maybe with decreasing proportion when the contact is older). 


```{r}
par(mar = rep(5, 4))
# Notifications per QR code, 1-day delay, raw
notifsPerQR.raw.decalage <- dat.all$diffNotifPerDay / (c(NA, dat.all$diffQRPerDay[-nrow(dat.all)]))

# Choice of the function over which cases are counted
prop.util <- function(x){
  sliding.fun(x, FUN = max, winwdt = 7, pos = 7)
}

plot(as.Date(dat.all$date), notifsPerQR.raw.decalage / prop.util(propUtilActif),       
     xlab = "", 
     ylab = "Notif. par QR code / max(Prop. utilisateurs 7 derniers j.)", 
     pch = 16, col = colActif, main = "", type = "n", 
     yaxs = "i", 
     frame.plot = FALSE)

title("Nb contacts détectés par utilisateur", line = 3)
mtext("(Estimation du nb notifs/QRcode si 100% utilisateurs)", line = 2, side = 3, cex = 0.8)
# Grid lines
# for(i in 0:65){
#   abline(h = i, col = gray(0.8))
# }
# for(i in seq(0, 65, by = 5)){
#   abline(h = i, col = gray(0.8), lwd = 2)
# }
for(i in seq(0, 65, by = 5)){
  abline(h = i, lty = 1, col = gray(0.8), lwd = 0.4)
}

axis(4)



# Annotate with dates
abline(v = as.Date(date_TAC), lty = 2)
cexAnnotate <- 0.8
mtext("StopCovid -> 
TousAntiCovid", side = 3, at = as.Date(date_TAC), adj = 1, cex = cexAnnotate)

abline(v = as.Date(date_sms), lty = 2)
mtext("envoi SMS pour 
QR code", side = 3, at = as.Date(date_sms), adj = 0, cex = cexAnnotate)

abline(v = as.Date(date_criteres), lty = 2)
mtext("nvx criteres
detection", side = 3, at = as.Date(date_criteres), adj = 0.7, cex = cexAnnotate)

abline(v = as.Date(date_criteres2), lty = 2)
mtext("nvx criteres
detection", side = 3, at = as.Date(date_criteres2), adj = 0.5, cex = cexAnnotate)


par(xpd = TRUE)
points(as.Date(dat.all$date), notifsPerQR.raw.decalage/prop.util(propUtilActif), pch = 16, col = colActif)

points(as.Date(dat.all$date), notifsPerQR.raw.decalage/prop.util(propUtil.Raw), col = colRegis, pch = 16)
par(xpd = FALSE)


legend("topleft", col = c("white", colActif, colRegis, "white"), pch = 16, 
       legend = c("prop. utilisateurs ", "Actifs", "Enregistrés"), 
       box.lwd = 0)


mtext("Date", side = 1, line = 1.5)

mtext(paste0("Mise à jour : ", Sys.Date(), "
Données : https://www.data.gouv.fr/fr/datasets/metriques-dutilisation-de-lapplication-tousanticovid/ pour TAC, 
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series 
pour nb cas en date d'annonce"), cex = 0.55, col = gray(0.3), side = 1, line = 3.25, adj = 0)

```

